Supplementary Information for\\ ``A weakly structured stem for human origins in Africa''

### 2023-03-16 supplement.tex
Address Tim's comments in supp

In this context, a number of possible models have been outlined
\citep{Groucutt2015-do}. Among these, \citeauthor{Groucutt2015-do} enumerate
variants on an early onset dispersal model. There is persuasive
paleo-morphological evidence for a Homo sapiens presence in Arabia between
120--85ka, coincident with Marine Isotope Stage (MIS) 5. The crania identified
at Skhul and Qafzeh in the Levant, as well a finger bone from southern Arabian
Peninsula point to a dispersal OOA when conditions in North Africa facilitated
a connection through the Sinai land bridge, and/or a possible connection via
crossing the Red Sea. Ecological conditions would have been similar between
North Africa and the Arabian Peninsula at this time, with greater abundance in
game during the higher rainfall MIS 5. While Skhul and Qafzeh are taxonomically
assigned to Homo sapiens, we note that their morphological diversity (i.e. a
mix of derived and archaic traits) is unique among contemporary human
populations. Under some formulations of the early onset dispersal model,
archaeologists posit a further expansion of humans into South Asia or as far as
southeastern Asia \citep{Groucutt2015-do}.

However, the replacement of Homo sapiens by Neanderthals in the Levant (e.g.
Kebara 2) complicates the persistence of modern humans in the Arabian Peninsula
after 85ka. The end of MIS 5 coincides with increasing aridity in the region
which would have rendered much the Peninsula uninhabitable, except for a
secluded southern refugia, decreasing the probability of migrations through the
Sinai land bridge \citep{Beyer2021-ym}. This barrier to human migration OOA
via the northern route likely remained until the Holocene (as modeled by
precipitation amount).  A southern route across the Red Sea becomes possible
during MIS 4 (70ka) until 30ka \citep{Beyer2021-ym}. Even during MIS 4, there
were few plausible routes out of the southern Arabian Peninsula, except along
the Gulf of Oman or gaps in the Arabian Desert through the Zagros Mountains.
Neanderthal remains dating to this period have been identified in the Zagros
Mountains, adding a further potential barrier to expansion
\citep{Yousefi2020-et,Heydari-Guran2021-xy}.

Thus, either an incipient OOA population must have expanded beyond the Near
East by 85ka and then given rise to today’s Eurasian populations or a
successful expansion would likely post-date 65ka. Morphological and
archaeological evidence for Homo sapiens in the Near East reappears by 50ka in
the Levant with the Initial Upper Paleolithic stone tool industry
\citep{Boaretto2021-oi} and a cranium from Manot Cave
\citep{Hershkovitz2015-lq}. There is also strong consensus that a widespread
human presence in Europe and Asia is dated to after 45ka. Archaeological
\citep[e.g.,][]{Fewlass2020-se}, morphological \citep[e.g.,][]{Hublin2020-nk},
and genetic data \citep[e.g.,][]{Prufer2021-ao} all indicate that about 45ka
\emph{Homo sapiens} had expanded into eastern Europe and within 5,000 years
rapidly colonized the entire span of southern Europe. This period further
coincides with the dates for the last Neanderthal presence in Europe at
approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly in Asia,
evidence of the Initial Upper Paleolithic (IUP) occurs 39-45ka as far east as Lake
Baikal. Characteristics of the IUP support a coherent, cultural package which
may have been transmitted by demic expansion \citep{Kuhn2014-kj,Zwyns2021-zv}.
By 45ka, there is additionally an unequivocal \emph{H. sapiens} femur from
southern Siberia which contains derived alleles shared with later Eurasian
populations \citep{Fu2014-am}. The Ust’-Ishim
specimen contains evidence of admixture with Neanderthals, at a level similar
to contemporary East Asian genomes. This supports estimates of Neanderthal
admixture occurring earlier than 45ka, estimated to be by 47-65ka as inferred
from Neanderthal tract lengths \citep{Sankararaman2012-uw}.

In summary, a sustained, rapid human expansion across Eurasia dates to just
prior to 45ka and emerges from a Near Eastern population. The time of
population divergence between eastern Africans and this Near Eastern group is
difficult to pinpoint. An exit OOA and persistence prior to 65ka appears
unlikely to have contributed to the ancestors of most Eurasians. By 50ka, a
Homo sapiens population occupied the Levant, subsequently admixed and replaced
Neanderthals, and is a likely source for the Eurasian populations. Hence, we
fixed the eastern African/Eurasian divergence at 50ka, though arguably dates
even earlier up to 65ka would be appropriate (see SI Section 7.2).

>>>>

In this context, a number of possible models have been outlined
\citep{Groucutt2015-do}. Among these, \citeauthor{Groucutt2015-do} enumerate
variants on an early onset dispersal model. There is persuasive
paleoanthropological evidence for a \emph{Homo sapiens} presence in Arabia
between 85--120ka, coincident with Marine Isotope Stage (MIS) 5. The crania
identified at Skhul and Qafzeh in the Levant, as well a finger bone from
southern Arabian Peninsula point to a dispersal OOA when conditions in North
Africa facilitated a connection through the Sinai land bridge, and/or a
possible connection via crossing the Red Sea. Ecological conditions would have
been similar between North Africa and the Arabian Peninsula at this time, with
greater abundance in game during the higher rainfall MIS 5. While Skhul and
Qafzeh are taxonomically assigned to \emph{Homo sapiens}, we note that they are
morphologically quite different from contemporary populations
\citep{Pearson2008-ok}. Under some formulations of the early onset dispersal
model, archaeologists posit a further expansion of humans into South Asia or as
far as southeastern Asia \citep{Groucutt2015-do}.

However, the replacement of \emph{Homo sapiens} by Neanderthals in the Levant
(e.g.  Kebara 2) complicates the persistence of modern humans in the Arabian
Peninsula after 85ka \citep{Valladas1987-bk}. The end of MIS 5 coincides with
increasing aridity in the region which would have rendered much the Peninsula
uninhabitable, except for a secluded southern refugia, decreasing the
probability of migrations through the Sinai land bridge \citep{Beyer2021-ym}.
This barrier to human migration OOA via the northern route likely remained
until the Holocene (as modeled by precipitation amount). A southern route
across the Red Sea becomes possible during MIS 4 (70ka) until 30ka
\citep{Beyer2021-ym}. Even during MIS 4, there were few plausible routes out of
the southern Arabian Peninsula, except along the Gulf of Oman or gaps in the
Arabian Desert through the Zagros Mountains. Neanderthal remains dating to this
period have been identified in the Zagros Mountains, adding a further potential
barrier to expansion \citep{Yousefi2020-et,Heydari-Guran2021-xy}.

Thus, either an incipient OOA population must have expanded beyond the Near
East by 85ka and then given rise to today’s Eurasian populations or a
successful expansion would likely post-date 65ka. Morphological and
archaeological evidence for \emph{Homo sapiens} in the Near East reappears by
50ka in the Levant with the Initial Upper Paleolithic stone tool industry
\citep{Boaretto2021-oi} and a cranium from Manot Cave
\citep{Hershkovitz2015-lq}. There is also strong consensus that a widespread
human presence in Europe and Asia is dated to after 45ka. Archaeological
\citep[e.g.,][]{Fewlass2020-se}, morphological \citep[e.g.,][]{Hublin2020-nk},
and genetic \citep[e.g.,][]{Prufer2021-ao} data all indicate that about 45ka
\emph{Homo sapiens} had expanded into eastern Europe and within 5,000 years
rapidly colonized the entire span of southern Europe. This period further
coincides with the dates for the last Neanderthal presence in Europe at
approximately 40ka \citep{Deviese2017-de,Hublin2017-uy}. Similarly in Asia,
evidence of the Initial Upper Paleolithic (IUP) occurs 39-45ka as far east as
Lake Baikal. Characteristics of the IUP support a coherent, cultural package
which may have been transmitted by demic expansion
\citep{Kuhn2014-kj,Zwyns2021-zv}.  By 45ka, there is additionally an
unequivocal \emph{Homo sapiens} femur from southern Siberia which contains
derived alleles shared with later Eurasian populations \citep{Fu2014-am}. The
Ust’-Ishim specimen contains evidence of admixture with Neanderthals, at a
level similar to contemporary East Asian genomes. This supports estimates of
Neanderthal admixture occurring earlier than 45ka, estimated to be by 47-65ka
as inferred from Neanderthal tract lengths \citep{Sankararaman2012-uw}.

In summary, a sustained, rapid human expansion across Eurasia dates to just
prior to 45ka and emerges from a Near Eastern population. The time of
population divergence between eastern Africans and this Near Eastern group is
difficult to pinpoint. An exit OOA and persistence prior to 65ka appears
unlikely to have contributed to the ancestors of most Eurasians. By 50ka, a
\emph{Homo sapiens} population occupied the Levant, subsequently admixed and
replaced Neanderthals, and is a likely source for the Eurasian populations.
Hence, we fixed the eastern African/Eurasian divergence at 50ka, though
arguably dates even earlier up to 65ka would be appropriate (see SI Section
7.2).
